---
title: "updatedmedianpanalysis"
format: html
editor: visual
---

packages:

```{r}
require(classInt)
require(curl)
require(dbscan)
require(downloader)
require(dplyr)
require(fasterize)
require(foreign)
require(fpc)
require(fs)
require(geojsonio)
require(ggplot2)
require(h3jsr)
require(h3r)
require(here)
require(janitor)
require(osmdata)
require(osmextract)
require(plotly)
require(raster)
require(readxl)
require(RColorBrewer)
require(sf)
require(sp)
require(spData)
require(spatstat)
require(spatstat.geom)
require(stars)
require(terra)
require(tidyterra)
require(tidyverse)
require(tmap)
require(tmaptools)
require(DT)
require(tidyr)
require(stringr)
require(ggrepel)
require(ggtext)
require(countrycode)
require(ggbump)
require(MetBrewer)
require(scales)
require(data.table)
require(dendextend)
require(gridExtra)
require(circlize)
require(rnaturalearth)
require(cartogram)
require(ggforce)
require(purrr)
require(viridis)
```

## Import Data

Waste water data:

```{r}
#Wste water EXPERIMENTAL DATA

    wwdata<- read.csv("wastewaterdata_denv_month_year_anonimized.csv")
    wwdata<- wwdata %>% filter(Region=="Portugal")
```

Flight data:

```{r}
## import flight dara

    flight <- readr::read_csv("mock_flight_data.csv",  locale = locale(encoding = "latin1"))
    flight <- janitor::clean_names(flight)

    as.data.frame(flight)
    medianflights <- flight

    #Organize and clean the district names.
    medianflights <- medianflights%>%
    mutate(
        District = case_when(
        dest %in% c("BEJA") ~ "Beja",
        dest %in% c("EVORA") ~ "Évora",
        dest %in% c("AVEIRO") ~ "Aveiro",
        dest %in% c("BRAGA") ~ "Braga",
        dest %in% c("BRAGANÇA") ~ "Bragança",
        dest %in% c("CASTELO BRANCO") ~ "Castelo Branco",
        dest %in% c("COIMBRA") ~ "Coimbra",
        dest %in% c("FARO", "PORTIMÃO") ~ "Faro",
        dest %in% c("GUARDA") ~ "Guarda",
        dest %in% c("Leiria") ~ "Leiria",
        dest %in% c("CASCAIS", "LISBOA") ~ "Lisboa",
        dest %in% c("FUNCHAL", "P.SANTO") ~ "Madeira",
        dest %in% c("PORTOALEGRE") ~ "Portalegre",
        dest %in% c("PORTO") ~ "Porto",
        dest %in% c("SANTAREM") ~ "Santarém",
        dest %in% c("SETUBAL") ~ "Setúbal",
        dest %in% c("VIANA DO CASTELO") ~ "Viana do Castelo",
        dest %in% c("VILA REAL") ~ "Vila Real",
        dest %in% c("VISEU") ~ "Viseu",
        dest %in% c("CORVO", "FLORES", "GRACIOSA", "HORTA", "LAJES", "P.DELGADA", "PICO", "S.JORGE", "S.MARIA") ~ "Azores",    
        TRUE ~ "Other"
        )
    )


    #remove azores and madeira
    
    medianflights <- medianflights %>%
        filter(!(dest %in% c("Azores", "Madeira")))


    
    # Fill in missing values with zeros

    medianflights <- medianflights %>%
        mutate(across(where(is.numeric), ~replace_na(., 0)))


    #Format as long

    medianflights <- medianflights %>%
    pivot_longer(cols = jan:dec, 
                names_to = "month", 
                values_to = "passengercount") %>%
    group_by(orig, month, year) %>%  # Group by country and year
    summarise(passenger_count = sum(passengercount, na.rm = TRUE), .groups = "drop") 



    # Convert month names to numbers
    months_num <- match(medianflights$month, c("jan", "feb", "mar", "apr", "may", "jun","jul", "aug", "sep", "oct", "nov", "dec"))


    # Create the new formatted date column
    medianflights <- medianflights %>%
    mutate(year = as.Date(paste(year, months_num, "02", sep = "-")))
    medianflights$year <- as.Date(medianflights$year, format = "%Y.%m.%d")


    #Translate country names to English
    medianflights <- medianflights%>%
    mutate(
        Origin = case_when(
        orig %in% c("Angola") ~ "Angola",
        orig %in% c("Brasil") ~ "Brazil",
        orig %in% c("China") ~ "China",
        orig %in% c("Colombia") ~ "Colombia",
        orig %in% c("Costa Rica") ~ "Costa Rica",
        orig %in% c("Cuba") ~ "Cuba",
        orig %in% c("República Dominicana") ~ "Dominican Republic",
        orig %in% c("França") ~ "France",
        orig %in% c("Guiana Francesa") ~ "frenchguiana",
        orig %in% c("Alemanha") ~ "Germany",
        orig %in% c("India") ~ "India",
        orig %in% c("Indonésia") ~ "Indonesia",
        orig %in% c("Maurício") ~ "Mauritius",
        orig %in% c("México") ~ "Mexico",
        orig %in% c("Moçambique") ~ "Mozambique",
        orig %in% c("Mianmar", "Birmânia") ~ "Myanmar",
        orig %in% c("Nepal") ~ "Nepal",
        orig %in% c("Nova Caledônia") ~ "newcaledonia",
        orig %in% c("Porto Rico") ~ "puertorico",
        orig %in% c("Portugal") ~ "Portugal",
        orig %in% c("Senegal") ~ "Senegal",
        orig %in% c("Tailândia") ~ "Thailand",
        orig %in% c("Venezuela") ~ "Venezuela",
        orig %in% c("Vietname") ~ "Vietnam",
        orig %in% c("Inglaterra") ~ "UnitedKingdom",
        TRUE ~ "Other"
        )
    )

    medianflights <- medianflights %>%
    filter(Origin != "Other") #Remove countries with other
    

    #remove month column
    medianflights <- medianflights %>% select(-contains("month")) 
    
    #remove orig columns
    medianflights <- medianflights %>% select(-c(orig))
    data.table::setnames(medianflights, "year", "date")
    data.table::setnames(medianflights, "Origin", "Country")
```

Index P Data:

```{r}
## INDEX P DATA

    #Create empty Data Frame for median data

    median_df <- data.frame(
        date = as.Date(character()),
        year = double(),
        month = character()
    )

    ##countries and regions
    #import each country or region and relabel data, then add to overall index p df

    Africa_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Africa_summaries_per_date.csv")
    Africa_medians <- Africa_medians %>% rename(Africa_median_p="median_p")
    median_df <- Africa_medians


    Americas_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Americas_summaries_per_date.csv")
    Americas_medians<- Americas_medians %>% rename(Americas_median_p="median_p")
    median_df <- median_df %>% left_join(Americas_medians, by = c("year", "month", "date"))

  

    Angola_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Angola_summaries_per_date.csv")
    Angola_medians<- Angola_medians %>% rename(Angola_median_p="median_p")
    median_df <- median_df %>% left_join(Angola_medians, by = c("year", "month", "date"))
    
    
    
    Asia_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Asia_summaries_per_date.csv")
    Asia_medians<- Asia_medians %>% rename(Asia_median_p="median_p")
    median_df <- median_df %>% left_join(Asia_medians, by = c("year", "month", "date"))


    Brazil_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Brazil_summaries_per_date.csv")
    Brazil_medians<- Brazil_medians %>% rename(Brazil_median_p="median_p")
    median_df <- median_df %>% left_join(Brazil_medians, by = c("year", "month", "date"))


    Caribbean_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Caribbean_summaries_per_date.csv")
    Caribbean_medians<- Caribbean_medians %>% rename(Caribbean_median_p="median_p")
    median_df <- median_df %>% left_join(Caribbean_medians, by = c("year", "month", "date"))
    

    
    CentralAmerica_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/CentralAmerica_summaries_per_date.csv")
    CentralAmerica_medians<- CentralAmerica_medians %>% rename(CentralAmerica_median_p="median_p")
    median_df <- median_df %>% left_join(CentralAmerica_medians, by = c("year", "month", "date"))
    
    
    China_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/China_summaries_per_date.csv")
    China_medians<- China_medians %>% rename(China_median_p="median_p")
    median_df <- median_df %>% left_join(China_medians, by = c("year", "month", "date"))
        

     Colombia_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Columbia_summaries_per_date.csv")
    Colombia_medians<- Colombia_medians %>% rename(Colombia_median_p="median_p")
    median_df <- median_df %>% left_join(Colombia_medians, by = c("year", "month", "date"))  

      CostaRica_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/CostaRica_summaries_per_date.csv")
    CostaRica_medians<- CostaRica_medians %>% rename(CostaRica_median_p="median_p")
    median_df <- median_df %>% left_join(CostaRica_medians, by = c("year", "month", "date"))
    
    
      Cuba_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Cuba_summaries_per_date.csv")
    Cuba_medians<- Cuba_medians %>% rename(Cuba_median_p="median_p")
    median_df <- median_df %>% left_join(Cuba_medians, by = c("year", "month", "date"))

    
      DominicanRepublic_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/DominicanRepublic_summaries_per_date.csv")
    DominicanRepublic_medians<- DominicanRepublic_medians %>% rename(DominicanRepublic_median_p="median_p")
    median_df <- median_df %>% left_join(DominicanRepublic_medians, by = c("year", "month", "date"))
    

    
      EastAfrica_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/EastAfrica_summaries_per_date.csv")
    EastAfrica_medians<- EastAfrica_medians %>% rename(EastAfrica_median_p="median_p")
    median_df <- median_df %>% left_join(EastAfrica_medians, by = c("year", "month", "date"))


      EastAsia_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/EastAsia_summaries_per_date.csv")
    EastAsia_medians<- EastAsia_medians %>% rename(EastAsia_median_p="median_p")
    median_df <- median_df %>% left_join(EastAsia_medians, by = c("year", "month", "date")) 
    

      Europe_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Europe_summaries_per_date.csv")
    Europe_medians<- Europe_medians %>% rename(Europe_median_p="median_p")
    median_df <- median_df %>% left_join(Europe_medians, by = c("year", "month", "date")) 
    
    
      France_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/France_summaries_per_date.csv")
   France_medians<- France_medians %>% rename(France_median_p="median_p")
    median_df <- median_df %>% left_join(France_medians, by = c("year", "month", "date")) 
    
    
      Germany_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Germany_summaries_per_date.csv")
   Germany_medians<- Germany_medians %>% rename(Germany_median_p="median_p")
    median_df <- median_df %>% left_join(Germany_medians, by = c("year", "month", "date")) 
        
    
      India_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/India_summaries_per_date.csv")
    India_medians<- India_medians %>% rename(India_median_p="median_p")
    median_df <- median_df %>% left_join(India_medians, by = c("year", "month", "date")) 
        
 
    
      Indonesia_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Indonesia_summaries_per_date.csv")
    Indonesia_medians<- Indonesia_medians %>% rename(Indonesia_median_p="median_p")
    median_df <- median_df %>% left_join(Indonesia_medians, by = c("year", "month", "date")) 



    
      Mauritius_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Mauritius_summaries_per_date.csv")
    Mauritius_medians<- Mauritius_medians %>% rename(Mauritius_median_p="median_p")
    median_df <- median_df %>% left_join(Mauritius_medians, by = c("year", "month", "date")) 
    
      Mexico_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Mexico_summaries_per_date.csv")
    Mexico_medians<- Mexico_medians %>% rename(Mexico_median_p="median_p")
    median_df <- median_df %>% left_join(Mexico_medians, by = c("year", "month", "date")) 
     
    

      MiddleAfrica_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/MiddleAfrica_summaries_per_date.csv")
    MiddleAfrica_medians<- MiddleAfrica_medians %>% rename(MiddleAfrica_median_p="median_p")
    median_df <- median_df %>% left_join(MiddleAfrica_medians, by = c("year", "month", "date")) 

    
    Mozambique_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Mozambique_summaries_per_date.csv")
    Mozambique_medians<- Mozambique_medians %>% rename(Mozambique_median_p="median_p")
    median_df <- median_df %>% left_join(Mozambique_medians, by = c("year", "month", "date"))

    
     Myanmar_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Myanmar_summaries_per_date.csv")
    Myanmar_medians<- Myanmar_medians %>% rename(Myanmar_median_p="median_p")
    median_df <- median_df %>% left_join(Myanmar_medians, by = c("year", "month", "date")) 
    
    Nepal_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Nepal_summaries_per_date.csv")
    Nepal_medians<- Nepal_medians %>% rename(Nepal_median_p="median_p")
    median_df <- median_df %>% left_join(Nepal_medians, by = c("year", "month", "date")) 
    
     Portugal_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Portugal_summaries_per_date.csv")
    Portugal_medians<- Portugal_medians %>% rename(Portugal_median_p="median_p")
    median_df <- median_df %>% left_join(Portugal_medians, by = c("year", "month", "date")) 
    
     Senegal_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Senegal_summaries_per_date.csv")
    Senegal_medians<- Senegal_medians %>% rename(Senegal_median_p="median_p")
    median_df <- median_df %>% left_join(Senegal_medians, by = c("year", "month", "date")) 
    

     SouthAmerica_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/SouthAmerica_summaries_per_date.csv")
    SouthAmerica_medians<- SouthAmerica_medians %>% rename(SouthAmerica_median_p="median_p")
    median_df <- median_df %>% left_join(SouthAmerica_medians, by = c("year", "month", "date")) 

    
     SouthAsia_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/SouthAsia_summaries_per_date.csv")
    SouthAsia_medians<- SouthAsia_medians %>% rename(SouthAsia_median_p="median_p")
    median_df <- median_df %>% left_join(SouthAsia_medians, by = c("year", "month", "date"))
    
    
     SoutheastAsia_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/SoutheastAsia_summaries_per_date.csv")
    SoutheastAsia_medians<- SoutheastAsia_medians %>% rename(SoutheastAsia_median_p="median_p")
    median_df <- median_df %>% left_join(SoutheastAsia_medians, by = c("year", "month", "date"))
    

     Thailand_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Thailand_summaries_per_date.csv")
    Thailand_medians<- Thailand_medians %>% rename(Thailand_median_p="median_p")
    median_df <- median_df %>% left_join(Thailand_medians, by = c("year", "month", "date"))

    
    
     UnitedKingdom_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/UnitedKingdom_summaries_per_date.csv")
    UnitedKingdom_medians<- UnitedKingdom_medians %>% rename(UnitedKingdom_median_p="median_p")
    median_df <- median_df %>% left_join(UnitedKingdom_medians, by = c("year", "month", "date"))


    
        
    Venezuela_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Venezuela_summaries_per_date.csv")
    Venezuela_medians<- Venezuela_medians %>% rename(Venezuela_median_p="median_p")
    median_df <- median_df %>% left_join(Venezuela_medians, by = c("year", "month", "date"))


    Vietnam_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/Vietnam_summaries_per_date.csv")
    Vietnam_medians<- Vietnam_medians %>% rename(Vietnam_median_p="median_p")
    median_df <- median_df %>% left_join(Vietnam_medians, by = c("year", "month", "date"))
    

    WestAfrica_medians <-  readr::read_csv("P_rasters_code/Index_P_summaries/WestAfrica_summaries_per_date.csv")
    WestAfrica_medians<- WestAfrica_medians %>% rename(WestAfrica_median_p="median_p")
    median_df <- median_df %>% left_join(WestAfrica_medians, by = c("year", "month", "date"))   
    
    
    #ensure it's a df
    median_df<- as.data.frame(median_df)
    
    #remove mean p as we will be working with medians
    median_df <- median_df %>% select(-contains("mean_p"))


    median_df_formated <- gather(median_df, key="Country", value="Transmission_Pot", -date, -month, -year)

    # Remove "_medians from column names" 
    median_df_formated <- median_df_formated %>% mutate(Country = gsub("_median_p", "", Country))
    
    #check head
    head(median_df_formated)

```

### Data Wrangling:

```{r}
## join index P and flight data

    #ensure data format
    median_df_formated$date <- as.Date(median_df_formated$date, format = "%Y/%m/%d")
    medianflights$date <- as.Date(medianflights$date, format = "%Y/%m/%d")
    median_df_formated <- median_df_formated %>%  mutate_if(is.Date, ~ . + 1)

    median_df_joined <- median_df_formated %>% left_join(medianflights, by = c("date" = "date", "Country" = "Country"))
    median_df_joined$passenger_count[is.na(median_df_joined$passenger_count)]<- 0
    
    
#Create regions and add passenger data to them    
    
#Africa:
        #Define countries    
        african_countries <- c("Angola", "Mauritius", "Mozambique", "Senegal")
        
        # Sum passenger_count per date for these countries
        africa_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% african_countries, ], 
                                 sum)
        
        # Append to "Africa"
        median_df_joined[median_df_joined$Country == "Africa", "passenger_count"] <- africa_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "Africa", "date"], africa_sums$date)]
#Americas:
         #Define countries    
        american_countries <- c("Brazil", "Colombia", "CostaRica", "Cuba", "Dominican Republic", "Mexico", "Venezuela")
        
        # Sum passenger_count per date for these countries
        americas_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% american_countries, ], 
                                 sum)
        
        # Append to "Americas" in its respective Country Column
        median_df_joined[median_df_joined$Country == "Americas", "passenger_count"] <- americas_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "Americas", "date"], americas_sums$date)]

#Asia:
        #Define countries    
        asian_countries <- c("China", "India", "Indonesia", "Myanmar", "Nepal", "Thailand", "Vietnam")
        
        # Sum passenger_count per date for these countries
        asia_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% asian_countries, ], 
                                 sum)
        
        # Append to "Asia" in its respective Country Column
        median_df_joined[median_df_joined$Country == "Asia", "passenger_count"] <- asia_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "Asia", "date"], asia_sums$date)]  
        
#Europe:
        #Define countries    
        european_countries <- c("France", "Germany", "UnitedKingdom")
        
        # Sum passenger_count per date for these countries
        europe_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% european_countries, ], 
                                 sum)
        
        # Append to "Europe" in its respective Country Column
        median_df_joined[median_df_joined$Country == "Europe", "passenger_count"] <- europe_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "Europe", "date"], europe_sums$date)]   

#Caribbean:
        #Define countries    
        caribbean_countries <- c("Cuba", "Dominican Republic")
        
        # Sum passenger_count per date for these countries
        caribbean_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% caribbean_countries, ], 
                                 sum)
        
        # Append to "caribbean" in its respective Country Column
        median_df_joined[median_df_joined$Country == "Caribbean", "passenger_count"] <- caribbean_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "Caribbean", "date"], caribbean_sums$date)]
        
#Central America:
        #Define countries    
        centralamerican_countries <- c("CostaRica", "Mexico")
        
        # Sum passenger_count per date for these countries
        centralamerican_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% centralamerican_countries, ], 
                                 sum)
        
        # Append to "centralamerican" in its respective Country Column
        median_df_joined[median_df_joined$Country == "CentralAmerica", "passenger_count"] <- centralamerican_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "CentralAmerica", "date"], centralamerican_sums$date)]
        
        
#East Africa:
        #Define countries    
        eastafrican_countries <- c("Mauritius", "Mozambique")
        
        # Sum passenger_count per date for these countries
        eastafrican_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% eastafrican_countries, ], 
                                 sum)
        
        # Append to "EastAfrica" in its respective Country Column
        median_df_joined[median_df_joined$Country == "EastAfrica", "passenger_count"] <- eastafrican_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "EastAfrica", "date"], eastafrican_sums$date)]
        
#East Asia:
        #Define countries    
        eastasian_countries <- c("China")
        
        # Sum passenger_count per date for these countries
        eastasian_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% eastasian_countries, ], 
                                 sum)
        
        # Append to "EastAsia" in its respective Country Column
        median_df_joined[median_df_joined$Country == "EastAsia", "passenger_count"] <- eastasian_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "EastAsia", "date"], eastasian_sums$date)]

        
#Middle Africa:
        #Define countries    
        midafrican_countries <- c("Angola")
        
        # Sum passenger_count per date for these countries
        midafrican_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% midafrican_countries, ], 
                                 sum)
        
        # Append to "MiddleAfrica" in its respective Country Column
        median_df_joined[median_df_joined$Country == "MiddleAfrica", "passenger_count"] <- midafrican_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "MiddleAfrica", "date"], midafrican_sums$date)]
        
#South America:
        #Define countries    
        southamerican_countries <- c("Brazil", "Colombia", "Venezuela")
        
        # Sum passenger_count per date for these countries
        southamerican_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% southamerican_countries, ], 
                                 sum)
        
        # Append to "SouthAmerican" in its respective Country Column
        median_df_joined[median_df_joined$Country == "SouthAmerica", "passenger_count"] <- southamerican_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "SouthAmerica", "date"], southamerican_sums$date)] 
        
#South Asia:
        #Define countries    
        southasian_countries <- c("India", "Myanmar","Nepal")
        
        # Sum passenger_count per date for these countries
        southasian_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% southasian_countries, ], 
                                 sum)
        
        # Append to "SouthAsia" in its respective Country Column
        median_df_joined[median_df_joined$Country == "SouthAsia", "passenger_count"] <- southasian_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "SouthAsia", "date"], southasian_sums$date)]
        

#Southeast Asia:
        #Define countries    
        southeastasian_countries <- c("Indonesia", "Thailand", "Vietnam")
        
        # Sum passenger_count per date for these countries
        southeastasian_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% southeastasian_countries, ], 
                                 sum)
        
        # Append to "SoutheastAsia" in its respective Country Column
        median_df_joined[median_df_joined$Country == "SoutheastAsia", "passenger_count"] <- southeastasian_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "SoutheastAsia", "date"], southeastasian_sums$date)]
        
#Western Africa:
        #Define countries    
        westafrican_countries <- c("Senegal")
        
        # Sum passenger_count per date for these countries
        westafrican_sums <- aggregate(passenger_count ~ date, 
                                 data = median_df_joined[median_df_joined$Country %in% westafrican_countries, ], 
                                 sum)
        
        # Append to "WestAfrica" in its respective Country Column
        median_df_joined[median_df_joined$Country == "WestAfrica", "passenger_count"] <- westafrican_sums$passenger_count[
          match(median_df_joined[median_df_joined$Country == "WestAfrica", "date"], westafrican_sums$date)]
        
        

    #Normalize by max
    median_df_joined$passenger_count_norm <- (median_df_joined$passenger_count/max(median_df_joined$passenger_count))
    median_df_joined$Transmission_Pot_norm <- (median_df_joined$Transmission_Pot/max(median_df_joined$Transmission_Pot))
```

```{r}

#check max transmission potential:
print(range(median_df_joined$Transmission_Pot_norm))

#check max passenger count:

print(range(median_df_joined$passenger_count))
```

Plot to check:

```{r}
ggplot(median_df_joined) + 
  geom_point(aes(x=Transmission_Pot, y=passenger_count)) +
  facet_wrap(~Country, scales="free")

ggplot(median_df_joined) + 
  geom_point(aes(x=Transmission_Pot_norm, y=passenger_count_norm)) +
  facet_wrap(~Country, scales="free")

ggplot(median_df_joined) + 
  geom_line(aes(x=date, y=passenger_count_norm)) +
  facet_wrap(~Country, scales="free")

```

Create Exclusion: Covid Years:

```{r}
median_df_joined_noCOVID<- median_df_joined %>% filter(! (date >= "2020-01-01" & date <= "2021-12-01"))

#plot check
ggplot(median_df_joined_noCOVID) + 
  geom_line(aes(x=date, y=passenger_count_norm)) +
  facet_wrap(~Country, scale = "free")
```

## Direct Risk Calculation - Simple Risk

```{r}

#set function
    do_actions_simplerisk<- function(country_label, country_data){

        #calculate median passenger count per country
        Kb<- (median(country_data$passenger_count[country_data$Country == country_label]))
        
        #multiply mean passenger by date index p
        risk_pmedian_co <- Kb*country_data$Transmission_Pot_norm

        return(risk_pmedian_co)
    }

    
    #select countries to use
    # all_countries<- c("Angola","Brazil", "China", "Colombia", "Costa_Rica", "Cuba", "Dominican Republic", "France", "frenchguiana", "Germany", "India", "Indonesia", "Mauritius", "Mexico", "Mozambique", "Myanmar", "Nepal", "portugal", "Senegal", "Thailand", "Venezuela","Vietnam", "UnitedKingdom")
    all_countries<- unique(median_df_joined_noCOVID$Country)

    #Set empty data frame
    results_simpleriskdf <- data.frame()

    # Loop over countries
    for (country_label in all_countries) {
        
      
  #select ensure dates are in correct format
  all_dates<-  range(as.Date(unique(median_df_joined_noCOVID$date), "%Y-%m-%d"))
  all_dates<-  seq(all_dates[1],all_dates[2],by="1 month")

    for (idate in 1:length(all_dates)){

      this_date<- all_dates[idate]
      print(paste0(country_label,"::",this_date))
            
  ## Filter per this i country
  country_data_median <- median_df_joined_noCOVID %>% filter(Country == country_label) %>% filter(date == this_date)

      if(nrow(country_data_median)==0){
        #add NA when this fails
        res<- data.frame(Country = country_label, risk_pmedian = NA, date = this_date)
            }else{
                    
        #apply method and bind country results
        results_simplerisk <- do_actions_simplerisk(country_label = country_label, country_data = country_data_median)
  
        res<- data.frame(Country = country_label, risk_pmedian = results_simplerisk, date = this_date)       
            }
            results_simpleriskdf <- rbind( results_simpleriskdf, res )
        }
    }

    head(results_simpleriskdf, n=100)
```

plot check:

```{r}
ggplot(results_simpleriskdf) +
  geom_line(aes(x=date, y=risk_pmedian)) + facet_wrap(~Country, scale = "free")
```

Summarize data by time frame:

```{r}
#### summarize per month

    risk_month_simple<- results_simpleriskdf
    risk_month_simple<- risk_month_simple %>% mutate(month=format(date,"%m"))
    risk_month_simple<- risk_month_simple %>% group_by(Country, month) %>% summarise(risk_sum=sum(risk_pmedian, na.rm = TRUE), risk_mean=mean(risk_pmedian, na.rm = TRUE), risk_med=median(risk_pmedian, na.rm = TRUE),  risk_sd=sd(risk_pmedian, na.rm = TRUE)) %>% data.frame() 
    
    #plot check
    ggplot(risk_month_simple) +
        geom_line(aes(x=month, y=risk_sum, group=1)) + facet_wrap(~Country, scales="free")
    
### summarize per year

    risk_year_simple<- results_simpleriskdf
    risk_year_simple<- risk_year_simple %>% mutate(year=format(date,"%Y"))
    risk_year_simple<- risk_year_simple %>% group_by(Country, year) %>% summarise(risk_sum=sum(risk_pmedian, na.rm = TRUE), risk_mean=mean(risk_pmedian, na.rm = TRUE), risk_med=median(risk_pmedian, na.rm = TRUE),  risk_sd=sd(risk_pmedian, na.rm = TRUE)) %>% data.frame() 
    
    
    #plot check
    ggplot(risk_year_simple) +
        geom_line(aes(x=year, y=risk_sum, group=1)) + facet_wrap(~Country, scales="free")

    
#Summarize overall    
  
    risk_overall_simple<- results_simpleriskdf
    risk_overall_simple<- risk_overall_simple %>% mutate(month=format(date,"%m"))
    risk_overall_simple<- risk_overall_simple %>% group_by(month) %>% summarise(risk_sum=sum(risk_pmedian, na.rm = TRUE), risk_mean=mean(risk_pmedian, na.rm = TRUE), risk_med=median(risk_pmedian, na.rm = TRUE),  risk_sd=sd(risk_pmedian, na.rm = TRUE)) %>% data.frame() 
    
    #plot check
    ggplot(risk_overall_simple) +
        geom_line(aes(x=month, y=risk_sum, group=1))
    
# Change year and month in respective DF to numeric        
risk_month_simple$month <- as.numeric(risk_month_simple$month)

risk_year_simple$year <- as.numeric(risk_year_simple$year)

risk_overall_simple$month <- as.numeric(risk_overall_simple$month)
```

```{r}
#Create df to check overall risk, no country risk
    risk_month_nocountry_simple<- results_simpleriskdf
    risk_month_nocountry_simple <- risk_month_nocountry_simple %>% mutate(month=format(date,"%m"))
    risk_month_nocountry_simple <- risk_month_nocountry_simple %>% group_by(month) %>% summarise(risk_sum=sum(risk_pmedian, na.rm = TRUE), risk_mean=mean(risk_pmedian, na.rm = TRUE), risk_med=median(risk_pmedian, na.rm = TRUE), risk_sd=sd(risk_pmedian, na.rm = TRUE)) %>%
      data.frame()

#plot
ggplot(risk_month_nocountry_simple) +
        geom_line(aes(x=month, y=risk_med-risk_sd, group=1), linetype=2) +
        geom_line(aes(x=month, y=risk_med+risk_sd, group=1), linetype=2) +
        geom_line(aes(x=month, y=risk_med, group=1)) 
```

### Plots:

#### Country Bump plots

##### Month

```{r}
#create ranking from month data and select countries
ranked8 <- risk_month_simple |> 
  filter(month %in% c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12) & !Country %in% c("Portugal", "France", "Germany", "Nepal", "Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica", "SouthAmerica", "WestAfrica", "SouthAsia", "median_p", "Caribbean", "SoutheastAsia")) |> 
  mutate(ranking = rank(desc(risk_sum), ties.method = "first"), .by = month) |> 
  mutate(ctry = Country)
geo <- ranked8$Country

selected1 <- c("Brazil", "Angola", "Cuba", "Thailand", "China", "UnitedKingdom")


#create plot
plotmonthcountrysimple <- ranked8 |>
  ggplot(aes(x = month, y = ranking, group = Country)) +
  geom_bump(linewidth = 0.6, color = "gray90", smooth = 6) +
  geom_bump(aes(color = Country), linewidth = 0.8, smooth = 6,
                    data = ~. |> filter(Country %in% selected1)) +
  geom_point(color = "white", size = 4) +
  geom_point(color = "gray90", size = 2) +
  geom_point(aes(color = Country), size = 2, 
             data = ~. |> filter(Country %in% selected1)) +
  geom_text(aes(label = ctry), x = 13, hjust = 0,
            color = "gray50", family = "Arial", size = 3.5,
            data = ranked8 |> slice_max(month, by = Country) |> 
              filter(!Country %in% selected1)) +
  geom_text(aes(label = ctry), x = 13, hjust = 0,
            color = "black", family = "Arial", size = 3.5,
            data = ranked8 |> slice_max(month, by = Country) |> 
              filter(Country %in% selected1)) +
  scale_color_manual(values = met.brewer("Juarez")) +
  scale_x_continuous(limits = c(0.9, 20), expand = c(0.01,0),
                     breaks =  c(1,2,3,4,5,6,7,8,9,10,11,12)) +
  scale_y_reverse(breaks = c(25,20,15,10,5,1), expand = c(0.02,0),
                  labels = number_format(suffix = ".")) +
  labs(x = NULL, y = NULL,
       title = toupper("Monthly Cumulative Risk - Linear Model"),
       subtitle = "",
       caption = "") +
  theme_minimal(base_family = "Arial", base_size = 12) +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 14, hjust = .5),
        plot.subtitle = element_text(size = 10, hjust = .5),
        plot.caption = element_text(size = 8))

#plot
plotmonthcountrysimple
```

##### Year

```{r}
#create ranking from month data and select countries
ranked8 <- risk_year_simple |> 
  filter(year %in% c(2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2021, 2022) & !Country %in% c("Portugal", "France", "Germany", "Nepal", "Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica", "SouthAmerica", "WestAfrica", "SouthAsia", "median_p", "Caribbean", "SoutheastAsia")) |> 
  mutate(ranking = rank(desc(risk_sum), ties.method = "first"), .by = year) |> 
  mutate(ctry = Country)
geo <- ranked8$Country

selected1 <- c("Brazil", "Angola", "Cuba", "Thailand", "China")

#Create plot
plotyearcountrysimple <- ranked8 |>
  ggplot(aes(x = year, y = ranking, group = Country)) +
  geom_bump(linewidth = 0.6, color = "gray90", smooth = 6) +
  geom_bump(aes(color = Country), linewidth = 0.8, smooth = 6,
                    data = ~. |> filter(Country %in% selected1)) +
  geom_point(color = "white", size = 4) +
  geom_point(color = "gray90", size = 2) +
  geom_point(aes(color = Country), size = 2, 
             data = ~. |> filter(Country %in% selected1)) +
  geom_text(aes(label = ctry), x = 2023, hjust = 0,
            color = "gray50", family = "Arial", size = 3.5,
            data = ranked8 |> slice_max(year, by = Country) |> 
              filter(!Country %in% selected1)) +
  geom_text(aes(label = ctry), x = 2023, hjust = 0,
            color = "black", family = "Arial", size = 3.5,
            data = ranked8 |> slice_max(year, by = Country) |> 
              filter(Country %in% selected1)) +
  scale_color_manual(values = met.brewer("Juarez")) +
  scale_x_continuous(limits = c(2011.9, 2027), expand = c(0.01,0),
                     breaks =  c(2012, 2017, 2019, 2021, 2022)) +
  scale_y_reverse(breaks = c(25,20,15,10,5,1), expand = c(0.02,0),
                  labels = number_format(suffix = ".")) +
  labs(x = NULL, y = NULL,
       title = toupper("Yearly Cumulative Risk by Country - Linear"),
       subtitle = "",
       caption = "") +
  theme_minimal(base_family = "Arial", base_size = 12) +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 14, hjust = .5),
        plot.subtitle = element_text(size = 10, hjust = .5),
        plot.caption = element_text(size = 8))

#plot
plotyearcountrysimple
```

#### Region Bump Plots

##### Month

```{r}

#create ranking from month data and select countries
ranked6 <- risk_month_simple |> 
  filter(month %in% c(1,2,3,4,5,6,7,8,9,10,11,12) & Country %in% c( "Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica", "SouthAmerica", "WestAfrica", "SouthAsia", "Caribbean", "SoutheastAsia")) |> 
  mutate(ranking = rank(desc(risk_sum), ties.method = "first"), .by = month) |> 
  mutate(ctry = Country)
geo <- ranked6$Country

#select countries to highlight
selected5 <- c("SouthAmerica", "Asia", "Africa", "Europe")

#create plot
plotmonthregionsimple <- ranked6 |>
  ggplot(aes(x = month, y = ranking, group = Country)) +
  geom_bump(linewidth = 0.6, color = "gray90", smooth = 6) +
  geom_bump(aes(color = Country), linewidth = 0.8, smooth = 6,
                    data = ~. |> filter(Country %in% selected5)) +
  geom_point(color = "white", size = 4) +
  geom_point(color = "gray90", size = 2) +
  geom_point(aes(color = Country), size = 2, 
             data = ~. |> filter(Country %in% selected5)) +
  geom_text(aes(label = ctry), x = 13, hjust = 0,
            color = "gray50", family = "Arial", size = 3.5,
            data = ranked6 |> slice_max(month, by = Country) |> 
              filter(!Country %in% selected5)) +
  geom_text(aes(label = ctry), x = 13, hjust = 0,
            color = "black", family = "Arial", size = 3.5,
            data = ranked6 |> slice_max(month, by = Country) |> 
              filter(Country %in% selected5)) +
  scale_color_manual(values = met.brewer("Juarez")) +
  scale_x_continuous(limits = c(0.9, 20), expand = c(0.01,0),
                     breaks =  c(1,2,3,4,5,6,7,8,9,10,11,12)) +
  scale_y_reverse(breaks = c(25,20,15,10,5,1), expand = c(0.02,0),
                  labels = number_format(suffix = ".")) +
  labs(x = NULL, y = NULL,
       title = toupper("Monthly Cumulative Risk by Region - Linear"),
       subtitle = "",
       caption = "") +
  theme_minimal(base_family = "Arial", base_size = 12) +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 14, hjust = .5),
        plot.subtitle = element_text(size = 10, hjust = .5),
        plot.caption = element_text(size = 8))

#plot
plotmonthregionsimple
```

##### Year

```{r}

#create yearly ranking and select countries/ years
ranked5 <- risk_year_simple |> 
  filter(year %in% c(2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2021, 2022) & Country %in% c( "Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica", "SouthAmerica", "WestAfrica", "SouthAsia", "Caribbean", "SoutheastAsia")) |> 
  mutate(ranking = rank(desc(risk_sum), ties.method = "first"), .by = year) |> 
  mutate(ctry = Country)
geo <- ranked5$Country

#see above chunk for selected5

#Create plot
plotyearregionsimple <- ranked5 |>
  ggplot(aes(x = year, y = ranking, group = Country)) +
  geom_bump(linewidth = 0.6, color = "gray90", smooth = 6) +
  geom_bump(aes(color = Country), linewidth = 0.8, smooth = 6,
                    data = ~. |> filter(Country %in% selected5)) +
  geom_point(color = "white", size = 4) +
  geom_point(color = "gray90", size = 2) +
  geom_point(aes(color = Country), size = 2, 
             data = ~. |> filter(Country %in% selected5)) +
  geom_text(aes(label = ctry), x = 2023, hjust = 0,
            color = "gray50", family = "Arial", size = 3.5,
            data = ranked5 |> slice_max(year, by = Country) |> 
              filter(!Country %in% selected5)) +
  geom_text(aes(label = ctry), x = 2023, hjust = 0,
            color = "black", family = "Arial", size = 3.5,
            data = ranked5 |> slice_max(year, by = Country) |> 
              filter(Country %in% selected5)) +
  scale_color_manual(values = met.brewer("Juarez")) +
  scale_x_continuous(limits = c(2011.9, 2027), expand = c(0.01,0),
                     breaks =  c(2012, 2017, 2019, 2021, 2022)) +
  scale_y_reverse(breaks = c(25,20,15,10,5,1), expand = c(0.02,0),
                  labels = number_format(suffix = ".")) +
  labs(x = NULL, y = NULL,
       title = toupper("Yearly Comulative Risk by Region"),
       subtitle = "",
       caption = "") +
  theme_minimal(base_family = "Arial", base_size = 12) +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 14, hjust = .5),
        plot.subtitle = element_text(size = 10, hjust = .5),
        plot.caption = element_text(size = 8))

#plot
plotyearregionsimple
```

##### Map:

```{r}
#load geo data
world <- ne_countries(scale = 110, type = "countries", returnclass = "sf")%>%
  # Convert WGS84 to projected crs
  sf::st_transform(world_ne, crs="ESRI:54030")


#bring in data
mapdata <- median_df_joined_noCOVID

#bring in risk data
data1 <- results_simpleriskdf


#join risk and flight data
mapdata1 <- mapdata %>%
  left_join(data1, by = c("date" = "date", "Country" = "Country"))

#remove regions from data
mapdata1 <- mapdata1 %>%
filter(!(Country %in% c("Portugal", "France", "Germany", "Nepal", "Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica","SouthAmerica", "WestAfrica", "SouthAsia", "median_p", "Caribbean", "SoutheastAsia")))

#Change country names to remove spaces so it matches geo data
mapdata1 <- mapdata1 %>%
  mutate(Country = ifelse(Country == "DominicanRepublic", "DominicanRepublic", Country))
mapdata1 <- mapdata1 %>%
  mutate(Country = ifelse(Country == "UnitedKingdom", "United Kingdom", Country))
mapdata1 <- mapdata1 %>%
  mutate(Country = ifelse(Country == "CostaRica", "Costa Rica", Country))


#Summarise data by country
mapdata1 <- mapdata1 %>% 
  group_by(Country) %>% 
  summarise(
    risk_sum = sum(risk_pmedian, na.rm = TRUE),
    risk_mean = mean(risk_pmedian, na.rm = TRUE),
    risk_med = median(risk_pmedian, na.rm = TRUE),
    risk_sd = sd(risk_pmedian, na.rm = TRUE),
    Transmission_Pot = mean(Transmission_Pot, na.rm = TRUE),  
    passenger_count = sum(passenger_count, na.rm = TRUE)  
  ) %>%
  data.frame()




# Join data to world map

map1 <- world %>%
  left_join(mapdata1, by = c("name" = "Country"))



```

Plot map:

```{r}
ggplot(map1, aes(fill = risk_sum)) +
  geom_sf(color = "white", size = 0.1) +
  scale_fill_viridis_c(
    option = "viridis",
    na.value = "lightgrey",
    trans = "log",
    name = "Log Cummulative Risk"
  ) +
  theme_minimal() +
  labs(title = "Cumulative Risk per Select Country - Linear Method")
```

## Entropy Risk Calc:

#### By Country

```{r}
  do_actions_risk<- function(country_label, country_data){

          
  #Normalize Index P s.t. sum equals 1:
  country_combined_df <- country_data %>% 
    mutate(pindexnorm_co =Transmission_Pot/ sum(Transmission_Pot))

  #Check norm worked:
  print(sum(country_combined_df$pindexnorm_co))
          
          
          
  #Calcuate H_Country

    country_combined_df <- country_data %>% mutate(hpi =  Transmission_Pot_norm^3 * log(Transmission_Pot_norm))
    country_Hc <- -sum(country_combined_df$hpi)
    print(range(country_Hc))

  #Calculate Risk:
  Overall_risk_co <- (country_Hc)*10^(4) * (median(country_data$passenger_count_norm[country_data$Country == country_label]))

            return(Overall_risk_co)
        }

  all_countries<- c("Angola","Brazil", "China", "Colombia", "CostaRica", "Cuba", "Dominican Republic", "India", "Indonesia", "Mauritius", "Mexico", "Mozambique", "Myanmar", "Senegal", "Thailand", "Venezuela","Vietnam", "UnitedKingdom")
        #all_countries<- unique(median_df_joined_noCOVID$Country)

    results_medianrisk_df <- data.frame()

        # Loop over countries
        for (country_label in all_countries) {

            this_country_df <- median_df_joined_noCOVID %>% filter(Country == country_label)
            # head(this_country_df)
            
            all_dates<-  as.Date(unique(median_df_joined_noCOVID$date), "%Y-%m-%d")

            for (idate in 1:length(all_dates)){

                this_date<- all_dates[idate]
                
                ## Filter this country
                this_year_country_data <- this_country_df %>% filter(date == this_date)
                head(this_year_country_data)
                        
                ##apply method
                results_median_riskdf <- do_actions_risk(country_label = country_label, country_data = this_year_country_data)
                            
                results_medianrisk_df <- rbind( results_medianrisk_df,
                                                    data.frame(Country = country_label, Overall_risk = results_median_riskdf, date = this_date))
            }
        }


        results_medianrisk_df <-  results_medianrisk_df %>% mutate(across(where(is.numeric), ~replace_na(., 0)))


```

Plot check:

```{r}
ggplot(results_medianrisk_df) +
            geom_line(aes(x=date, y=Overall_risk)) + facet_wrap(~Country, scale = "free")

```

Summarize data per time frame:

```{r}
#summarize per month

    risk_month<- results_medianrisk_df
    risk_month<- risk_month %>% mutate(month=format(date,"%m"))
    risk_month<- risk_month %>% group_by(Country, month) %>% summarise(risk_sum=sum(Overall_risk), risk_mean=mean(Overall_risk), risk_med=median(Overall_risk),  risk_sd=sd(Overall_risk)) %>% data.frame() 
    
  #plot check
  ggplot(risk_month) +
    geom_line(aes(x=month, y=risk_sum, group=1)) + facet_wrap(~Country, scales="free")
    
#summarize per year

    risk_year<- results_medianrisk_df
    risk_year<- risk_year %>% mutate(year=format(date,"%y"))
    risk_year<- risk_year %>% group_by(Country, year) %>% summarise(risk_sum=sum(Overall_risk), risk_mean=mean(Overall_risk), risk_med=median(Overall_risk),  risk_sd=sd(Overall_risk)) %>% data.frame() 
    
        #plot check
    ggplot(risk_year) +
        geom_line(aes(x=year, y=risk_sum, group=1)) + facet_wrap(~Country, scales="free")

#Sumarize Overall:
    
    risk_over<- results_medianrisk_df
    risk_over<- risk_over %>% mutate(month=format(date,"%m"))
    risk_over<- risk_over %>% group_by(month) %>% summarise(risk_sum=sum(Overall_risk), risk_mean=mean(Overall_risk), risk_med=median(Overall_risk),  risk_sd=sd(Overall_risk)) %>% data.frame() 
    
    
  #plot check
  ggplot(risk_over) +
    geom_line(aes(x=month, y=risk_sum, group=1))




```

```{r}
#create df to check risk overall, no country
    risk_month_nocountry<- results_medianrisk_df
    risk_month_nocountry<- risk_month_nocountry %>% mutate(month=format(date,"%m"))
    risk_month_nocountry<- risk_month_nocountry %>% group_by(month) %>% summarise(risk_sum=sum(Overall_risk), risk_mean=mean(Overall_risk), risk_med=median(Overall_risk), risk_sd=sd(Overall_risk)) %>%  data.frame()

#plot
ggplot(risk_month_nocountry) +
        geom_line(aes(x=month, y=risk_med-risk_sd, group=1), linetype=2) +
        geom_line(aes(x=month, y=risk_med+risk_sd, group=1), linetype=2) +
        geom_line(aes(x=month, y=risk_med, group=1)) 
```

Risk Calc by region:

```{r}
  do_actions_risk<- function(country_label, country_data){

    #Normalize Index P s.t. sum equals 1:
    country_combined_df <- country_data %>% 
      mutate(pindexnorm_co =Transmission_Pot/ sum(Transmission_Pot))

    #Check norm worked:
    print(sum(country_combined_df$pindexnorm_co))
          
          
          
  #Calcuate H_Country

  country_combined_df <- country_data %>% mutate(hpi =  Transmission_Pot_norm^3 * log(Transmission_Pot_norm))
  country_Hc <- -sum(country_combined_df$hpi)
  # print(country_Hc)

  #Calculate Risk:
  Overall_risk_co <- (country_Hc)*10^(4) * (median(country_data$passenger_count_norm[country_data$Country == country_label]))

            return(Overall_risk_co)
        }

         all_countries1<- c("Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica", "SouthAmerica", "WestAfrica", "SouthAsia", "Caribbean", "SoutheastAsia")
        #all_countries<- unique(median_df_joined_noCOVID$Country)

        results_regionmedian_riskdf <- data.frame()

        # Loop over countries
        for (country_label in all_countries1) {



            this_country_df <- median_df_joined_noCOVID %>% filter(Country == country_label)
            # head(this_country_df)
            
            all_dates<-  as.Date(unique(median_df_joined_noCOVID$date), "%Y-%m-%d")

            for (idate in 1:length(all_dates)){

                this_date<- all_dates[idate]
                
            #Filter this country
            this_year_country_data <- this_country_df %>% filter(date == this_date)
            head(this_year_country_data)
                        
                #apply method
                results_regionmedian_risk <- do_actions_risk(country_label = country_label, country_data = this_year_country_data)
                            
                results_regionmedian_riskdf <- rbind( results_regionmedian_riskdf,
                                                    data.frame(Country = country_label, Overall_risk = results_regionmedian_risk, date = this_date))
            }
        }


results_regionmedian_riskdf <-  results_regionmedian_riskdf %>% mutate(across(where(is.numeric), ~replace_na(., 0)))

```

Join country and region:

```{r}

entropyjoined <- bind_rows(results_medianrisk_df, results_regionmedian_riskdf)


#summarize by month:

risk_month_all_entropy<- entropyjoined
risk_month_all_entropy<- risk_month_all_entropy %>% mutate(month=format(date,"%m"))
risk_month_all_entropy<- risk_month_all_entropy %>% group_by(Country, month) %>% summarise(risk_sum=sum(Overall_risk), risk_mean=mean(Overall_risk), risk_med=median(Overall_risk),  risk_sd=sd(Overall_risk)) %>% data.frame() 
```

Summarize by time frame:

```{r}
#### summarize per month

  risk_month_region<- results_regionmedian_riskdf
  risk_month_region<- risk_month_region %>% mutate(month=format(date,"%m"))
  risk_month_region<- risk_month_region %>% group_by(Country, month) %>% summarise(risk_sum=sum(Overall_risk), risk_mean=mean(Overall_risk), risk_med=median(Overall_risk),  risk_sd=sd(Overall_risk)) %>% data.frame() 
    
  #plot check
  ggplot(risk_month_region) +
    geom_line(aes(x=month, y=risk_sum, group=1)) + facet_wrap(~Country, scales="free")
    
### summarize per year


  risk_year_region<- results_regionmedian_riskdf
  risk_year_region<- risk_year_region %>% mutate(year=format(date,"%y"))
  risk_year_region<- risk_year_region %>% group_by(Country, year) %>% summarise(risk_sum=sum(Overall_risk), risk_mean=mean(Overall_risk), risk_med=median(Overall_risk),  risk_sd=sd(Overall_risk)) %>% data.frame() 
    
    
  #plot check
  ggplot(risk_year_region) +
    geom_line(aes(x=year, y=risk_sum, group=1)) + facet_wrap(~Country, scales="free")

```

### Plots

#### Monthly Plots:

```{r}
#change risk_month$month to numerica

risk_month$month <- as.numeric(risk_month$month)

risk_month_region$month <- as.numeric(risk_month_region$month)
```

Country Bump plots:

```{r}
ranked1 <- risk_month |> 
  filter(month %in% c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12) & !Country %in% c("Portugal", "France", "Germany", "Nepal", "Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica", "SouthAmerica", "WestAfrica", "SouthAsia", "median_p", "Caribbean", "SoutheastAsia")) |> 
  mutate(ranking = rank(desc(risk_sum), ties.method = "first"), .by = month) |> 
  mutate(ctry = Country)
geo <- ranked1$Country

selected1 <- c("Brazil", "Angola", "Cuba", "Thailand", "China", "UnitedKingdom")
```

```{r}
plotmonthlycountry <- ranked1 |>
  ggplot(aes(x = month, y = ranking, group = Country)) +
  geom_bump(linewidth = 0.6, color = "gray90", smooth = 6) +
  geom_bump(aes(color = Country), linewidth = 0.8, smooth = 6,
                    data = ~. |> filter(Country %in% selected1)) +
  geom_point(color = "white", size = 4) +
  geom_point(color = "gray90", size = 2) +
  geom_point(aes(color = Country), size = 2, 
             data = ~. |> filter(Country %in% selected1)) +
  geom_text(aes(label = ctry), x = 13, hjust = 0,
            color = "gray50", family = "Arial", size = 3.5,
            data = ranked1 |> slice_max(month, by = Country) |> 
              filter(!Country %in% selected1)) +
  geom_text(aes(label = ctry), x = 13, hjust = 0,
            color = "black", family = "Arial", size = 3.5,
            data = ranked1 |> slice_max(month, by = Country) |> 
              filter(Country %in% selected1)) +
  scale_color_manual(values = met.brewer("Juarez")) +
  scale_x_continuous(limits = c(0, 20), expand = c(0.01,0),
                     breaks = c(1, 3, 8, 12)) +
  scale_y_reverse(breaks = c(25,20,15,10,5,1), expand = c(0.02,0),
                  labels = number_format(suffix = ".")) +
  labs(x = NULL, y = NULL,
       title = toupper("Monthly Cumulative Risk - Weighted Entropy Model"),
       subtitle = "",
       caption = "tbd") +
  theme_minimal(base_family = "Arial", base_size = 12) +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 14, hjust = .5),
        plot.subtitle = element_text(size = 10, hjust = .5),
        plot.caption = element_text(size = 8))

plotmonthlycountry
```

Monthly Region Plots:

```{r}
ranked2 <- risk_month_region |> 
  filter(month %in% c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12) & Country %in% c( "Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica", "SouthAmerica", "WestAfrica", "SouthAsia", "Caribbean", "SoutheastAsia")) |> 
  mutate(ranking = rank(desc(risk_sum), ties.method = "first"), .by = month) |> 
  mutate(ctry = Country)
geo <- ranked2$Country

selected2 <- c("SouthAmerica", "Asia", "Africa", "Europe")


plotmonthlyregion <- ranked2 |>
  ggplot(aes(x = month, y = ranking, group = Country)) +
  geom_bump(linewidth = 0.6, color = "gray90", smooth = 6) +
  geom_bump(aes(color = Country), linewidth = 0.8, smooth = 6,
                    data = ~. |> filter(Country %in% selected2)) +
  geom_point(color = "white", size = 4) +
  geom_point(color = "gray90", size = 2) +
  geom_point(aes(color = Country), size = 2, 
             data = ~. |> filter(Country %in% selected2)) +
  geom_text(aes(label = ctry), x = 13, hjust = 0,
            color = "gray50", family = "Arial", size = 3.5,
            data = ranked2 |> slice_max(month, by = Country) |> 
              filter(!Country %in% selected2)) +
  geom_text(aes(label = ctry), x = 13, hjust = 0,
            color = "black", family = "Arial", size = 3.5,
            data = ranked2 |> slice_max(month, by = Country) |> 
              filter(Country %in% selected2)) +
  scale_color_manual(values = met.brewer("Juarez")) +
  scale_x_continuous(limits = c(0, 20), expand = c(0.01,0),
                     breaks = c(1, 3, 8, 12)) +
  scale_y_reverse(breaks = c(25,20,15,10,5,1), expand = c(0.02,0),
                  labels = number_format(suffix = ".")) +
  labs(x = NULL, y = NULL,
       title = toupper("Monthly Overall Risk by Region - Logarithm"),
       subtitle = "",
       caption = "") +
  theme_minimal(base_family = "Arial", base_size = 12) +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 14, hjust = .5),
        plot.subtitle = element_text(size = 10, hjust = .5),
        plot.caption = element_text(size = 8))

plotmonthlyregion
```

#### Yearly plots:

By country plot:

```{r}

#reformat year as factor

risk_year$year <- as.numeric(risk_year$year)

risk_year_region$year <- as.numeric(risk_year_region$year)

#create ranked list:
ranked3 <- risk_year |> 
  filter(year %in% c(12, 13, 14, 15, 16, 17, 18, 19, 21, 22) & !Country %in% c("Portugal", "France", "Germany", "Nepal", "Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica", "SouthAmerica", "WestAfrica", "SouthAsia", "median_p", "Caribbean", "SoutheastAsia")) |> 
  mutate(ranking = rank(desc(risk_sum), ties.method = "first"), .by = year) |> 
  mutate(ctry = Country)
geo <- ranked3$Country

#select highlighted countries:
selected3 <- c("Brazil", "Angola", "Cuba", "Thailand", "China")
```

```{r}
#plot bump plot:
plotyearcountry <- ranked3 |>
  ggplot(aes(x = year, y = ranking, group = Country)) +
  geom_bump(linewidth = 0.6, color = "gray90", smooth = 6) +
  geom_bump(aes(color = Country), linewidth = 0.8, smooth = 6,
                    data = ~. |> filter(Country %in% selected3)) +
  geom_point(color = "white", size = 4) +
  geom_point(color = "gray90", size = 2) +
  geom_point(aes(color = Country), size = 2, 
             data = ~. |> filter(Country %in% selected3)) +
  geom_text(aes(label = ctry), x = 23, hjust = 0,
            color = "gray50", family = "Arial", size = 3.5,
            data = ranked3 |> slice_max(year, by = Country) |> 
              filter(!Country %in% selected3)) +
  geom_text(aes(label = ctry), x = 23, hjust = 0,
            color = "black", family = "Arial", size = 3.5,
            data = ranked3 |> slice_max(year, by = Country) |> 
              filter(Country %in% selected3)) +
  scale_color_manual(values = met.brewer("Juarez")) +
  scale_x_continuous(limits = c(11.9, 27), expand = c(0.01,0),
                     breaks = c(12, 17, 19, 21, 22)) +
  scale_y_reverse(breaks = c(25,20,15,10,5,1), expand = c(0.02,0),
                  labels = number_format(suffix = ".")) +
  labs(x = NULL, y = NULL,
       title = toupper("Yearly Overall Risk by Country - Entropy "),
       subtitle = "",
       caption = "") +
  theme_minimal(base_family = "Arial", base_size = 12) +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 14, hjust = .5),
        plot.subtitle = element_text(size = 10, hjust = .5),
        plot.caption = element_text(size = 8))

plotyearcountry
```

By region bump plot:

```{r}
ranked4 <- risk_year_region |> 
  filter(year %in% c(12, 13, 14, 15, 16, 17, 18, 19, 21, 22) & Country %in% c( "Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica", "SouthAmerica", "WestAfrica", "SouthAsia", "Caribbean", "SoutheastAsia")) |> 
  mutate(ranking = rank(desc(risk_sum), ties.method = "first"), .by = year) |> 
  mutate(ctry = Country)
geo <- ranked1$Country

#set highlighted countries
selected4 <- c("SouthAmerica", "Asia", "Africa", "Europe")

#plot
plotyearregion <- ranked4 |>
  ggplot(aes(x = year, y = ranking, group = Country)) +
  geom_bump(linewidth = 0.6, color = "gray90", smooth = 6) +
  geom_bump(aes(color = Country), linewidth = 0.8, smooth = 6,
                    data = ~. |> filter(Country %in% selected4)) +
  geom_point(color = "white", size = 4) +
  geom_point(color = "gray90", size = 2) +
  geom_point(aes(color = Country), size = 2, 
             data = ~. |> filter(Country %in% selected4)) +
  geom_text(aes(label = ctry), x = 23, hjust = 0,
            color = "gray50", family = "Arial", size = 3.5,
            data = ranked4 |> slice_max(year, by = Country) |> 
              filter(!Country %in% selected4)) +
  geom_text(aes(label = ctry), x = 23, hjust = 0,
            color = "black", family = "Arial", size = 3.5,
            data = ranked4 |> slice_max(year, by = Country) |> 
              filter(Country %in% selected4)) +
  scale_color_manual(values = met.brewer("Juarez")) +
  scale_x_continuous(limits = c(11.9, 27), expand = c(0.01,0),
                     breaks =  c(12, 17, 19, 21, 22)) +
  scale_y_reverse(breaks = c(25,20,15,10,5,1), expand = c(0.02,0),
                  labels = number_format(suffix = ".")) +
  labs(x = NULL, y = NULL,
       title = toupper("Yearly Comulative Risk by Country"),
       subtitle = "",
       caption = "") +
  theme_minimal(base_family = "Arial", base_size = 12) +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 14, hjust = .5),
        plot.subtitle = element_text(size = 10, hjust = .5),
        plot.caption = element_text(size = 8))

plotyearregion 
```

#### Bubble plot:

```{r}

#set df for bubbleplot data
bubbleplotdata <- median_df_joined_noCOVID

data <-results_regionmedian_riskdf 

#join dfs with overall and risk data
bubbleplotdata <- bubbleplotdata %>%
  left_join(data, by = c("date" = "date", "Country" = "Country"))

#remove regions
bubbleplotdata <- bubbleplotdata %>%
  filter(Country %in% c("Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica", "SouthAmerica", "WestAfrica", "SouthAsia", "median_p", "Caribbean", "SoutheastAsia"))


#set data by reporting metric
bubbleplotdata <- bubbleplotdata %>% 
  group_by(Country, year) %>% 
  summarise(
    risk_sum = sum(Overall_risk, na.rm = TRUE),
    risk_mean = mean(Overall_risk, na.rm = TRUE),
    risk_med = median(Overall_risk, na.rm = TRUE),
    risk_sd = sd(Overall_risk, na.rm = TRUE),
    Transmission_Pot = mean(Transmission_Pot, na.rm = TRUE),  # Aggregating by mean (or change to sum, etc.)
    passenger_count = sum(passenger_count, na.rm = TRUE)  # Aggregating by sum (or change to mean, etc.)
  ) %>%
  data.frame()

#set country column
bubbleplotdata$Country1 <- bubbleplotdata$Country

bubbleplotdata$Country <- paste(bubbleplotdata$Country,bubbleplotdata$year)
```

```{r}

#Save plot as png
png("bubbleplot.png", width = 1600, height = 1600, res = 150)

#plot
bubbleplotdata %>%
  arrange(desc(risk_sum)) %>%
  mutate(Country = factor(Country)) %>%
  ggplot(aes(x = passenger_count, 
             y = Transmission_Pot, 
             size = risk_sum, 
             color = Country)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(0.1, 24), name = "Yearly Cumulative Risk") +
  scale_x_continuous(trans = "log10")+
  scale_color_discrete(guide = "none") +  
  facet_wrap(~Country1, scales = "free")+
   labs(y = "Baseline Index P", x = "Baseline Passenger Counts")
dev.off()

```

#### Map:

```{r}


#load geo data
world <- ne_countries(scale = 110, type = "countries", returnclass = "sf")%>%
  # Convert WGS84 to projected crs
  sf::st_transform(world_ne, crs="ESRI:54030")


#bring in flight data
mapdata <- median_df_joined_noCOVID

#bring in entropy risk data
data <- results_medianrisk_df


#join risk and flight data
mapdata <- mapdata %>%
  left_join(data, by = c("date" = "date", "Country" = "Country"))

#remove regions from data
mapdata <- mapdata %>%
filter(!(Country %in% c("Portugal", "France", "Germany", "Nepal", "Africa", "Americas", "Europe", "CentralAmerica", "EastAfrica", "EastAsia", "Asia", "MiddleAfrica","SouthAmerica", "WestAfrica", "SouthAsia", "median_p", "Caribbean", "SoutheastAsia")))

#Change country names to remove spaces so it matches geo data
mapdata <- mapdata %>%
  mutate(Country = ifelse(Country == "DominicanRepublic", "DominicanRepublic", Country))
mapdata <- mapdata %>%
  mutate(Country = ifelse(Country == "UnitedKingdom", "United Kingdom", Country))
mapdata <- mapdata %>%
  mutate(Country = ifelse(Country == "CostaRica", "Costa Rica", Country))


#Summarise data by country
mapdata <- mapdata %>% 
  group_by(Country) %>% 
  summarise(
    risk_sum = sum(Overall_risk, na.rm = TRUE),
    risk_mean = mean(Overall_risk, na.rm = TRUE),
    risk_med = median(Overall_risk, na.rm = TRUE),
    risk_sd = sd(Overall_risk, na.rm = TRUE),
    Transmission_Pot = mean(Transmission_Pot, na.rm = TRUE),  
    passenger_count = sum(passenger_count, na.rm = TRUE)  
  ) %>%
  data.frame()




# Join data to world map

map <- world %>%
  left_join(mapdata, by = c("name" = "Country"))



```

```{r}

#plot map
ggplot(map, aes(fill = risk_sum)) +
  geom_sf(color = "white", size = 0.1) +
  scale_fill_viridis_c(
    option = "viridis",
    na.value = "lightgrey",
    trans = "log",
    name = "Log Cummulative Risk"
  ) +
  theme_minimal() +
  labs(title = "Overall Risk per Select Country - Weighted-entropy Method")
```

## Wastewater Analysis

```{r}
# extract waste water period 

wwdata_month <- wwdata %>% mutate(month=as.numeric(format(as.Date(month_year,"%Y-%m-%d"),"%m"))) %>% select(-month_year)
    
ggplot(wwdata_month) +
  geom_point(aes(x=month, y=DENV))+
    theme_minimal() +
    labs(title = "Dengue Wastewater RNA Data")
```

Entropy:

```{r}

#Merge risk data and wastewater data:

risk_ww_month<- wwdata_month %>% left_join(risk_month_all_entropy %>% mutate(month=as.numeric(month)), by="month") 

#plot wastewater DENV RNA findings vs risk summaries:  

g1<- ggplot(risk_ww_month) +
        geom_point(aes(x=risk_sum, y=log10(1+DENV), group=1)) + facet_wrap(~Country, scales="free")

g2<- ggplot(risk_ww_month) +
        geom_point(aes(x=risk_mean, y=log10(1+DENV), group=1)) + facet_wrap(~Country, scales="free")

g3<- ggplot(risk_ww_month) +
        geom_point(aes(x=risk_med, y=log10(1+DENV), group=1)) + facet_wrap(~Country, scales="free")

g2
g1
g3


```

```{r}
# extract waste water period overall risk

    ww_data <- wwdata %>% rename(date="month_year") %>% mutate(date=as.Date(date,"%Y-%m-%d")-13)
    risk_ww <- ww_data %>% left_join(results_medianrisk_df, by="date") %>% filter(!is.na(Overall_risk))

    ggplot(risk_ww) +
        geom_point(aes(x=Overall_risk, y=log10(1+DENV), group=1)) + facet_wrap(~Country, scales="free")
```

Statistical testing:

```{r}

#Run correlation testing: spearman

corr_results <- map_dfr(unique(risk_ww_month$Country), function(this_country) {
  df_subset <- filter(risk_ww_month, Country == this_country)
  test_result <- cor.test(log10(1 + df_subset$DENV), df_subset$risk_med, method = "spearman")
  
  tibble(
    Country = this_country,
    Spearman_corr = test_result$estimate,
    p_value = test_result$p.value,
    Significance = ifelse(test_result$p.value < 0.05, "Significant", "Not Significant")
    
  )
})



#Plot spearman results

logcorr <- ggplot(corr_results, aes(y = Country, x = Spearman_corr)) +
  geom_point(size = 3) +
  geom_text(aes(label = round(Spearman_corr, 2)), vjust = -0.5, color = "black") +
  labs(
    title = "Spearman's Correlation by Country Weighted Entropy Method",
    x = "Correlation Coefficient",
    y = "Country/ Region"
  ) +
  scale_color_manual(values = c("Significant" = "blue", "Not Significant" = "gray")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(t = 0, b = 20, r = 20, l = 20), 
    plot.title = element_text(size = 12, hjust = 0, vjust = 1)  
  ) +
  coord_cartesian(clip = "off")+
  geom_vline(xintercept = 0, linetype = "dashed", color = "red")

logcorr
```

Overall:

```{r}
#Merge risk data and wastewater data:
risk_ww_overall <- wwdata_month %>% left_join(risk_over %>% mutate(month=as.numeric(month)), by="month") 
#plot wastewater DENV RNA findings vs risk summeries:  

g13<- ggplot(risk_ww_overall) +
        geom_point(aes(x=risk_sum, y=log10(1+DENV))) 

g14<- ggplot(risk_ww_overall) +
        geom_point(aes(x=risk_mean, y=log10(1+DENV))) 

g15<- ggplot(risk_ww_overall) +
        geom_point(aes(x=risk_med, y=log10(1+DENV))) 

g13
g14
g15
```

```{r}
# Spearman correlation test
spearman_result <- cor.test(risk_ww_overall$risk_sum, risk_ww_overall$DENV, method = "spearman")
```

```{r}
#plot spearman
overalllogspearplot <- ggplot(risk_ww_overall, aes(x = month)) +
  geom_line(aes(y = risk_sum, color = "Overall Risk"), size = 1) +
  geom_point(aes(y = (DENV)/2000, color = "Wastewater RNA"), size = 4, shape = 21, fill = "navyblue") +
  labs(
    title = "Monthly Weighted Entropy Model Risk vs. Wastewater RNA Levels",
    subtitle = paste0("Spearman ρ = ", round(spearman_result$estimate, 2),
                      ", p = ", signif(spearman_result$p.value, 2)),
    x = "Month",
    y = "DENV/2000",
    color = "Legend"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

overalllogspearplot
```

Direct:

```{r}

#Merge risk data and wastewater data:
risk_ww_month_simple <- wwdata_month %>% left_join(risk_month_simple %>% mutate(month=as.numeric(month)), by="month") 
#plot wastewater DENV RNA findings vs risk summeries:  

g4<- ggplot(risk_ww_month_simple) +
        geom_point(aes(x=risk_sum, y=log10(1+DENV), group=1)) + facet_wrap(~Country, scales="free")

g5<- ggplot(risk_ww_month_simple) +
        geom_point(aes(x=risk_mean, y=log10(1+DENV), group=1)) + facet_wrap(~Country, scales="free")

g6<- ggplot(risk_ww_month_simple) +
        geom_point(aes(x=risk_med, y=log10(1+DENV), group=1)) + facet_wrap(~Country, scales="free")

g4
g5
g6
```

```{r}
# extract waste water period overall risk

    ww_data_simple <- wwdata %>% rename(date="month_year") %>% mutate(date=as.Date(date,"%Y-%m-%d")-13)
    risk_ww_simple <- ww_data_simple %>% left_join(results_simpleriskdf, by="date") %>% filter(!is.na(risk_pmedian))

    ggplot(risk_ww_simple) +
        geom_point(aes(x=risk_pmedian, y=log10(1+DENV), group=1)) + facet_wrap(~Country, scales="free")
```

Run correlation analysis:

```{r}
#Run correlation testing: spearman

corr_results_simple <- map_dfr(unique(risk_ww_month_simple$Country), function(this_country) {
  df_subset_simple <- filter(risk_ww_month_simple, Country == this_country)
  test_result_simple <- cor.test(log10(1 + df_subset_simple$DENV), df_subset_simple$risk_med, method = "spearman")
  
  tibble(
    Country = this_country,
    Spearman_corr = test_result_simple$estimate,
    p_value = test_result_simple$p.value
  )
})

#Plot spearman results

simplecorr <- ggplot(corr_results_simple, aes(y = Country, x = Spearman_corr)) +
  geom_point(size = 3) +
  geom_text(aes(label = round(Spearman_corr, 2)), vjust = -0.5, color = "black") +
  labs(
    title = "Spearman's Correlation by Country Linear Method",
    x = "Correlation Coefficient",
    y = "Country/ Region"
  ) +
  scale_color_manual(values = c("Significant" = "blue", "Not Significant" = "gray")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  coord_cartesian(clip = "off")+
  geom_vline(xintercept = 0, linetype = "dashed", color = "red")

simplecorr
```

Overall monthly risk:

```{r}
#Merge risk data and wastewater data:
risk_ww_overall_simple <- wwdata_month %>% left_join(risk_overall_simple %>% mutate(month=as.numeric(month)), by="month") 
#plot wastewater DENV RNA findings vs risk summeries:  

g10<- ggplot(risk_ww_overall_simple) +
        geom_point(aes(x=risk_sum, y=log10(1+DENV), group=1)) 

g11<- ggplot(risk_ww_month_simple) +
        geom_point(aes(x=risk_mean, y=log10(1+DENV), group=1)) 

g12<- ggplot(risk_ww_month_simple) +
        geom_point(aes(x=risk_med, y=log10(1+DENV), group=1)) 

g10
g11
g12
```

```{r}
# Spearman correlation test
spearman_result <- cor.test(risk_ww_overall_simple$risk_sum, risk_ww_overall_simple$DENV, method = "spearman")


```

```{r}
#plot spearman
overalllinearspearplot <- ggplot(risk_ww_overall_simple, aes(x = month)) +
  geom_line(aes(y = risk_sum, color = "Cumultive Risk"), size = 1) +
  geom_point(aes(y = DENV, color = "Wastewater RNA"), size = 4, shape = 24, fill = "navyblue") +
  labs(
    title = "Monthly Linear Model Risk vs. Wastewater RNA Levels",
    subtitle = paste0("Spearman ρ = ", round(spearman_result$estimate, 2),
                      ", p = ", signif(spearman_result$p.value, 2)),
    x = "Month",
    y = "DENV",
    colour = "Legend"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

overalllinearspearplot
```

## Sensitivity Analysis:

### Direct Risk:

```{r}
#Create df of theoretical index p and passenger count ranges for sensitivity analysis:

transmission_pot_seq <- seq(0, 1, by = 0.01)
passenger_count_seq <- seq(0, 1204203, by = 50000)



rangeanalysis_df <- expand.grid(
  Transmission_Pot = transmission_pot_seq,
  passenger_count = passenger_count_seq
)

#Normalize passenger count data:
rangeanalysis_df$passenger_count <-( rangeanalysis_df$passenger_count/ (max(rangeanalysis_df$passenger_count)))

#format date column
rangeanalysis_df$date <- as.Date("2020-01-01")

#Give name to our made-up country
rangeanalysis_df$Country <- "Genovia"
```

```{r}
 #Normalize by max
    rangeanalysis_df$passenger_count <- ( rangeanalysis_df$passenger_count/max( rangeanalysis_df$passenger_count))
    rangeanalysis_df$pindexnorm_gv <- (rangeanalysis_df$Transmission_Pot/max(rangeanalysis_df$Transmission_Pot))

#Check norm worked:
print(sum(rangeanalysis_df$pindexnorm_gv))


#Calculate Risk:

    #Passenger count influence
    Kb<- (mean(rangeanalysis_df$passenger_count))
            
    risk_pmedian_co <- rangeanalysis_df$passenger_count*rangeanalysis_df$pindexnorm_gv
  
    rangeanalysis_df <- rangeanalysis_df %>%
      mutate(ithrisk =  abs(rangeanalysis_df$passenger_count * rangeanalysis_df$pindexnorm_gv *1))

Overall_risk_gv <- abs(rangeanalysis_df$passenger_count* rangeanalysis_df$pindexnorm_gv)

print(Overall_risk_gv)
```

Heat plot:

```{r}
# Ensure risk is numeric
rangeanalysis_df$ithrisk <- as.numeric(rangeanalysis_df$ithrisk)

# Create binned copy
rangeanalysis_df_bin <- rangeanalysis_df
rangeanalysis_df_bin$ithrisk <- cut(rangeanalysis_df_bin$ithrisk, breaks = 7)

# Make plot
linearsensitivity <- ggplot(rangeanalysis_df_bin, aes(x = passenger_count, y = Transmission_Pot, fill = ithrisk)) +
  geom_tile() +
  labs(
    title = "Linear Model",
    x = "Passenger Count (in Mil.)", 
    y = "Index P", 
    fill = "Risk"
  ) +
  scale_fill_viridis_d(option = "A", direction = -1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

linearsensitivity
```

### Entropy-based risk:

```{r}

rangeanalysis_df_entropy <- rangeanalysis_df

#Normalize Index P s.t. sum equals 1:
rangeanalysis_df_entropy <- rangeanalysis_df_entropy %>%
   mutate(pindexnorm_gv = rangeanalysis_df_entropy$Transmission_Pot/ sum(rangeanalysis_df_entropy$Transmission_Pot))


#Check norm worked:
print(sum(rangeanalysis_df_entropy$pindexnorm))


#Calcuate H_country

rangeanalysis_df_entropy <- rangeanalysis_df_entropy %>%
  mutate(hpi =  abs(pindexnorm_gv^3 * log(pindexnorm_gv+0.00000000001)))

Genovia_Hc <- -sum(rangeanalysis_df_entropy$hpi)





#Calculate Risk:

rangeanalysis_df_entropy <- rangeanalysis_df_entropy %>%
  mutate(ithrisk =  (rangeanalysis_df_entropy$hpi * (rangeanalysis_df_entropy$passenger_count)))


```

Heat map of sensitivity analysis:

```{r}


```

```{r}

# Ensure risk is numeric
rangeanalysis_df_entropy$ithrisk <- as.numeric(rangeanalysis_df_entropy$ithrisk)

# Create a binned copy
rangeanalysis_df_entropy_bin <- rangeanalysis_df_entropy
rangeanalysis_df_entropy_bin$ithrisk <- cut(rangeanalysis_df_entropy_bin$ithrisk, breaks = 7)

#Heat map of sensitivity analysis:

entropysensitivity <- ggplot(rangeanalysis_df_entropy_bin, aes(x = passenger_count, y = Transmission_Pot, fill = ithrisk)) +
  geom_tile() +
  labs(
    title = "Weighted Entropy Model",
    x = "Passenger Count (in Mil.)", 
    y = "Index P", 
    fill = "Risk"
  ) +
  scale_fill_viridis_d(option = "A", direction = -1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

entropysensitivity
```

Extract diagonals for both methods:

```{r}
#make df with values in the diagonal where passenger count = index p for entropy model

diagonal_df_entropy <- rangeanalysis_df_entropy %>%
  filter(passenger_count == Transmission_Pot) %>%
  arrange(Transmission_Pot)

#remove hpi column to make it the same size as the linear one
diagonal_df_entropy <- diagonal_df_entropy %>% select(-hpi)


#create df with values in the diagonal for linear model
diagonal_df_linear <- rangeanalysis_df %>%
  filter(passenger_count == Transmission_Pot) %>%
  arrange(Transmission_Pot)



# Plot the diagonal line
ggplot(diagonal_df_entropy, aes(x = Transmission_Pot, y = (ithrisk))) +
  geom_line() +
  labs(title = "Risk along diagonal (Transmission Potential = Normalized Passenger Count)",
       x = "Index P = Travel",
       y = "Risk") +
  theme_minimal()


```

```{r}

#create column with model type
diagonal_df_entropy$model <- "entropy"

diagonal_df_linear$model <- "linear"

#joing both diagonal dfs previously made

combined_df <- rbind(diagonal_df_entropy, diagonal_df_linear)

#create plot

ggplot(combined_df, aes(x = Transmission_Pot, y = ithrisk)) +
  geom_line(color = "navyblue", size = 1) +
  facet_wrap(~model, scales = "free_y") +
  labs(title = "Risk along Diagonal (Index P = Air Passenger Count)",
       x = "Index P = Travel (normalized)",
       y = "Risk") +
  theme_minimal()
```

## Write-up graphs:

Initial code from some graphs from R Graph Gallery later modified for this project.

Create joined graph of yearly bump plots by country by both methods:

```{r}

#Save plot as png
png("yearlycountrybumps.png", width = 2000, height = 800, res = 150)

grid.arrange(plotyearcountry, plotyearcountrysimple, ncol = 2)

dev.off()

```

Create joined graph of monthly bump plots by country by both methods:

```{r}

#Save plot as png
png("monthlycountrybumps.png", width = 2000, height = 800, res = 150)

grid.arrange(plotmonthlycountry,plotmonthcountrysimple, ncol = 2)

dev.off()
```

Create joined graph of sensitivity analysis:

```{r}

#Save plot as png
png("sensitivity.png", width = 2000, height = 800, res = 150)

grid.arrange(entropysensitivity, linearsensitivity, ncol = 2)

dev.off()
```

Create joined graph of Spearman Analysis:

```{r}
#Save plot as png
png("validation.png", width = 2000, height = 800, res = 150)

grid.arrange(logcorr, simplecorr, ncol = 2)

dev.off()
```

Create joined graph of Overall Spearman Analysis:

```{r}
#Save plot as png
png("overallvalidation.png", width = 2000, height = 1000, res = 150)

grid.arrange(overalllogspearplot, overalllinearspearplot, ncol = 2)

dev.off()
```

Plot of modified entropy vs regular between \[0,1\]

```{r}

#set range for x
x <- seq(0.01, 1, length.out = 1000)

#deifne df
df <- data.frame(
  x = x,
  Original = -x * log(x),
  Weighted = -x^3 * log(x)
)

# format long
df_long <- pivot_longer(df, cols = c("Original", "Weighted"),
                        names_to = "Function", values_to = "y")

# Plot
ggplot(df_long, aes(x = x, y = y, color = Function)) +
  geom_line(size = 1) +
  geom_vline(xintercept = 0.38, linetype = "dashed", color = "navy") +
  geom_vline(xintercept = 0.716531, linetype = "dashed", color = "maroon") +
  annotate("text", x = 0.38, y = 0.3, label = "Maxima ~ 0.38", hjust = -0.1, color = "navy") +
  annotate("text", x = 0.716531, y = 0.3, label = "Maxima ~ 0.71", hjust = -0.1, color = "maroon") +
  scale_color_manual(values = c("Original" = "navy", "Weighted" = "maroon")) +
  labs(
    title = "Entropy Curve Comparison: Original vs Exponent-Weighted",
    x = "ith Index P value",
    y = "ith risk"
  ) +
  theme_minimal(base_size = 12)

```

```{r}
#plot Venezuela Index P as a check

#plot(y = Venezuela_medians$Venezuela_median_p, x=Venezuela_medians$date)

ggplot(Venezuela_medians, aes(x = date, y = Venezuela_median_p))+
  geom_line()+
  theme_minimal(base_size = 12)
  
```

Dendrogram:

```{r}
# Create a dendrogram:

 

dendata <- median_df_joined

dendata <- dendata %>% filter(!Country %in% c("portugal","Germany","France"))

dendata <- dendata %>%
  group_by(Country)%>%
  summarise(Median_Transmission_Pot = median(Transmission_Pot),
            passenger_count= sum(passenger_count))
  

# Clusterisation using 3 variables
row.names(dendata)=dendata$Country



dendata %>% 
  select(Median_Transmission_Pot) %>% as.data.frame()%>%
  dist() %>% 
  hclust() %>% 
  as.dendrogram() -> dend
labels(dend) <- row.names(dendata)
# Plot
par(mar=c(7,3,1,1))  # Increase bottom margin to have the complete label
plot(dend)
```

```{r}
# Color in function of the cluster
par(mar=c(1,1,1,7))
dend %>%
  set("labels_col", value = c("skyblue", "orange", "grey"), k=3) %>%
  set("branches_k_color", value = c("skyblue", "orange", "grey"), k = 3) %>%
  plot(horiz=TRUE, axes=FALSE)
abline(v = 350, lty = 2)
```

```{r}
row.names(dendata)=dendata$Country
dendata %>% 
  select(passenger_count) %>% as.data.frame()%>%
  dist() %>% 
  hclust(method = "complete") %>% 
  as.dendrogram() -> d2
labels(d2) <- row.names(dendata)
# Plot
par(mar=c(7,3,1,1)) 
plot(d2)
```

```{r}
dl <- dendlist(
  dend %>% 
    set("labels_col", value = c("skyblue", "orange", "grey"), k=3) %>%
    set("branches_lty", 1) %>%
    set("branches_k_color", value = c("skyblue", "orange", "grey"), k = 3),
  d2 %>% 
    set("labels_col", value = c("blue", "red", "green", "maroon", "violet"), k=5) %>%
    set("branches_lty", 1) %>%
    set("branches_k_color", value = c("blue", "red", "green", "maroon", "violet"), k = 5)
)
 
# Plot them together
tanglegram(dl, 
           common_subtrees_color_lines = FALSE, highlight_distinct_edges  = TRUE, highlight_branches_lwd=FALSE, 
           margin_inner=7,
           lwd=2
)

title(main="Dendrogram of Transmission Pot. vs Passenger", cex.main=0.8)
```
